Live Functional Programming with Typed Holes

### 2018-04-16 calculus.tex
near-final intro

The syntax of the core calculus, specified in Fig.~\ref{fig:hazelnut-live-syntax}, consists of types and expressions with holes. 
We distinguish between \emph{external} expressions, $e$, and \emph{internal} expressions, $d$. 
External expressions correspond to programs as entered by the programmer 
(see Sec.~\ref{sec:intro}\todo{or maybe Sec. 2?}{} for discussion of manual, semi-automated and fully automated hole entry methods). 
Each well-typed external expression (as specified in Sec.~\ref{sec:external-statics} below) expands to a well-typed internal expression (see Sec.~\ref{sec:expansion}) before it is evaluated (see Sec.~\ref{sec:evaluation}). 
We take this approach, notably also taken in the specification of Standard ML by \citet{Harper00atype-theoretic}, because (1) the external language supports type inference and explicit type ascriptions, $\hexp : \htau$, but it is formally simpler to eliminate ascriptions and specify a type assignment system when defining the dynamic semantics; and 
(2) we need additional syntactic machinery during evaluation for tracking hole closures and dynamic type casts. 
This machinery is inserted by the expansion step, rather than entered explicitly by the programmer. 
In this regard, the internal language is analogous to the cast calculus in the gradually typed lambda calculus \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will see the \HazelnutLive internal language goes beyond the cast calculus in several respects. We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently} 
(see Sec.~\ref{sec:agda-mechanization}). We have also implemented \HazelnutLive in a manner that maintains a novel end-to-end well-definedness invariant (see Sec.~\ref{sec:implementation}).

@@

We start with the type system of the \HazelnutLive external language,
which closely follows the \Hazelnut type system \cite{popl-paper},
except that (1) holes in \HazelnutLive each have a \emph{unqiue name},
and (2) for brievity of exposition, we consider a more austere core
language here that lacks numbers and sum types, which we discuss
(along with other systematic extensions) in
Appendix~\ref{sec:extensions}.

\Figref{fig:bidirectional-typing} defines the type system in the \emph{bidirectional} style 
%
with two mutually defined judgements \cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da}. The type synthesis
judgement~$\hsyn{\hGamma}{\hexp}{\htau}$ synthesizes a type~$\htau$
for external expression~$\hexp$ under typing context $\hGamma$, which tracks typing
assumptions of the form $x : \htau$ in the usual
manner \cite{pfpl,tapl}.
%
The type analysis judgement~$\hana{\hGamma}{\hexp}{\htau}$ checks
expression~$\hexp$ against a given type~$\htau$.
%
Algorithmically, analysis accepts a type as input, and synthesis gives
a type as output.
%
We start with synthesis for the programmer's ``top level'' external
expression.

@@

The primary benefit of specifying the \HazelnutLive external language 
bidirectionally is that the programmer need not annotate each hole with a type. 
%
An empty hole is
written $\hehole{u}$, where $u$ is a name for the hole, which we tacitly assume is unique.
%
%(Holes in \Hazelnut are \emph{unnamed}).
%
Rule {SEHole}\todo{rule name macros}{} specifies that an empty hole synthesizes hole type, written $\tehole$.
%
If an empty hole appears where an expression of some other type is
expected, e.g. under an explicit ascription (governed by Rule {SAsc})
or in the argument position of a function application (governed by
Rule {SAp}, discussed below), we apply the \emph{subsumption rule},
Rule {ASubsume}, which specifies that if an expression $e$ synthesizes
type $\htau$, then it may be checked against any \emph{consistent}
type, $\htau'$.

@@

For the familiar forms of the lambda calculus, the rules again follow prior work.
%
For simplicity, the core calculus includes only a single base type~$b$ with a single constant~$c$, governed by Rule {SConst} (i.e. $b$ is the unit type).
%
%\matt{Extraneous and uninteresting:}
%By contrast, \Hazelnut instead defines a number type with a single operation, which we include in Appendix~\ref{sec:extensions} alongside various other standard extensions to the core calculus\todo{do this, say more?}. 
%
Rule {SVar} specifies that variables synthesize the corresponding type from $\hGamma$.

@@

%The reason analytic expansion produces an expansion of consistent
%type is because the subsumption rule, as previously discussed, allows
%us to check an external expression against any type consistent with
%the type the expression actually synthesizes, whereas every internal
%expression can be assigned at most one type, i.e. the following
%standard unicity property holds of the type assignment system.
%
Analytic expansion produces an expansion of consistent type:
%
Intuitively, the subsumption rule, as previously discussed, allows us
to check an external expression against any type consistent with the
type the expression actually synthesizes.

@@

The rules governing expansion of constants, variables and lambda expressions
--- Rules {ESConst}, {ESVar}, {ESLam} and {EALam}
--- and the corresponding type assignment rules
--- Rules {TAConst}, {TAVar} and {TALam}
--- mirror the typing rules from Fig.~\ref{fig:bidirectional-typing}.
%
Consequently, the corresponding cases of
Theorem~\ref{thm:typed-expansion}, Theorem~\ref{thm:expandability} and
Theorem~\ref{thm:expansion-generality} are straightforward.
%
To support type assignment, all lambdas in the internal language are
half-annotated---Rule {EALam} inserts the annotation when expanding an
unannotated external lambda based on the given type.
%
Some rules perform \emph{cast insertion}, making them more interesting
than the others; these rules include those for holes, and the those for
function applications and ascriptions.  We consider each below.

\subsubsection{Hole Expansion}\label{sec:hole-expansion} 
Rules {ESEHole}, {ESNEHole}, {EAEHole} and {EANEHole} govern the expansion of empty and non-empty expression holes to empty and non-empty \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$, respectively. 
%
The hole name~$u$ on a hole closure identifies the external hole to which the hole closure corresponds.
%
%Note that
While we assume each hole name to be unique in the external language, due to substitution, evaluated (internal) terms may consist of multiple hole closures with the same name.
%
For example, the result from Fig.~\ref{fig:grades-example} shows four closures for the hole named 1\todo{update with actual number (four?) and hole name later}.
%
There, we number each hole closure for a given hole sequentially, \li{1:1}, \li{1:2} and so on, but this is strictly for the sake of presentation, so we omit hole closure numbers from the core calculus.

The hole expansion rules introduce hypotheses of the form~$\Dbinding{u}{\hGamma}{\htau}$ into the hole context~$\Delta$.
%
For each hole name~$u$, the hole context records a type~$\tau$, and a typing context~$\Gamma$.\footnote{
We use a hole context, rather than recording the typing context and type directly on each hole closure, to ensure that all closures for a hole name have the same typing context and type.}
%
We borrow this hole context notation from contextual modal type theory (CMTT) \cite{Nanevski2008}, identifying hole names with metavariables and hole contexts with modal contexts (we say more about the connection with CMTT below).
%
Each hole expansion rule records the ``current'' typing context under which the hole is expanded.
%
In the synthetic hole expansion rules~{ESEHole} and~{ESNEHole}, the generated hole context assigns the hole type~$\tehole$ to hole name~$u$, as in the typing rules.
%
However, the first two premises of the expansion subsumption rule~{EASubsume} disallow the use of subsumption for holes in analytic position.
%
Instead, we employ separate analytic rules~{EAEHole} and~{EANEHole}, which each record the checked type~$\tau$ in the hole context.
%
Consequently, we can use type assignment for the internal language --- the type assignment rules TAEHole and TANEHole in Fig.~\ref{fig:hasType} assign a hole closure for hole name~$u$ the corresponding type from the hole context.

Each hole closure also has an associated environment~$\sigma$ which consists of a finite substitution of the form $[d_1/x_1, ~\cdots, d_n/x_n]$ for $n \geq 0$.
%
The closure environment keeps a record of the substitutions that occur around the hole as evaluation occurs.
%
Initially, when no evaluation has yet occurred, the hole expansion rules generate an initial environment that consists of the identity substitution for the typing context associated with hole name~$u$ in hole context~$\Delta$, which we notate $\idof{\hGamma}$, and define as follows.
%
\begin{defn}[Identity Substitution] $\idof{x_1 : \tau_1, ~\cdots, x_n : \tau_n} = [x_1/x_1, ~\cdots, x_n/x_n]$
\end{defn}
\noindent
The type assignment rules for hole closures~{TAEHole} and {TANEHole} each require that the hole closure environment~$\sigma$ be consistent with the corresponding typing context, written as $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$.
%
Formally, we define this relation as follows:\todo{check that the definition in the Agda corresponds}
\begin{defn}[Substitution Typing]
$\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff $\domof{\sigma} = \domof{\hGamma'}$ and for each $x : \htau \in \hGamma'$ we have that $d/x \in \sigma$ and $\hasType{\Delta}{\hGamma}{d}{\tau}$.
\end{defn}
\noindent
It is easy to verify that the identity substitution satisfies this requirement, i.e. that $\hasType{\Delta}{\hGamma}{\idof{\hGamma}}{\hGamma}$.

Empty hole closures, $\dehole{u}{\sigma}{}$,  correspond to the metavariable closures (a.k.a. deferred substitutions) from CMTT, $\cmttclo{u}{\sigma}$.\todo{change notation from CMTT}{}
%
\Secref{sec:evaluation} defines how these closure environments evolve during evaluation.
%
Non-empty hole closures~$\dhole{d}{u}{\sigma}{}$ have no direct correspondence with a notion from CMTT (see Sec.~\ref{sec:resumption}).

\subsubsection{Cast Insertion}\label{sec:cast-insertion}
%
% USE A TOPIC SENTENCE SO THAT THE READER IS GUIDED A LITTLE MORE, e.g.,
%
Cast insertion mediates between the type systems of the external and
internal languages, where the former permits subsumption, and the
latter does not.
%-----------------------------------------------------------------------------------------------
%
To see why this mediation is necessary, consider the following
example: $\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$.
%
Viewed as an external expression, this example synthesizes type
$\tehole$, since the hole type annotation on variable~$x$ permits
applying~$x$ as a function of type~$\tarr{\tehole}{\tehole}$, and base
constant~$c$ may be checked against type~$\tehole$, by subsumption.
%
However, viewed as an internal expression, this example is not
well-typed---the type assignment system defined in
\Figref{fig:hasType} lacks subsumption.
%
Indeed, it would violate type safety if we could assign a type to this
example in the internal language, because beta reduction of this
example viewed as an internal expression would result in $c(c)$, which
is clearly not well-typed.
%
The difficulty arises because leaving the argument type unknown also leaves unknown how
the argument is being used (in this case, as a function).\footnote{In a system where type reconstruction is first used
to try to fill in type holes, we could express a similar example by
using $x$ at two or more different types, thereby causing type
reconstruction to fail.
%
% On the other hand, if it is acceptable to arbitrarily choose one of
%the possible types, and type reconstruction is complete, then type
%holes will never appear in the internal language and the cast
%insertion machinery described in this section can be omitted
%entirely, leaving only the hole closure machinery described
%previously.
}
\todo{talk about elsewhere? maybe do type-hole-free version of calculus in appendix if there is time}{} 
By our interpretation of hole types as unknown types from gradual type
theory, we can address the problem by performing cast insertion.
%

The cast form in Hazelnut Live consists of $\dcasttwo{\dexp}{\htau_1}{\htau_2}$.
%
This form serves to ``box'' an expression of type $\htau_1$ for
treatment as an expression of a consistent type $\htau_2$
(Rule TACast in \Figref{fig:hasType}).
%
\footnote{
In the earliest work on gradual type theory, the cast form only gave
the target type~$\htau_2$ \cite{Siek06a}, but it simplifies matters
to include the assigned type~$\htau_1$ in the syntax \cite{DBLP:conf/snapl/SiekVCB15}.
}

Expansion inserts casts for function applications and ascriptions.
%
The latter is more straightforward: Rule~{ESAsc}
in \Figref{fig:expandSyn} inserts a cast from the assigned type to the
ascribed type.
%
Theorem~\ref{thm:typed-expansion} inductively ensures that the two
types are consistent.
%
We include ascription for exposition purposes---this form is derivable
by using application together with the half-annotated identity, $e
: \tau = \hap{(\halam{x}{\htau}{x})}{e}$; as such, application
expansion, discussed below, is more general.

Rule~{ESAp} expands function applications.
%
To understand the rule, consider the expansion of external
expression~$\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$, the example
discussed above:
\[
        \hap{\dcasttwo{
        (\halam{x}{\tehole}{\underbrace{
                \hap{\dcasttwo{x}{\tehole}{\tarr{\tehole}{\tehole}}}
                {\dcasttwo{c}{b}{\tehole}}
                }_{\textrm{expansion of function body}~x(c)}
        }
        )}{\tarr{\tehole}{\tehole}}{
           \tarr{\tehole}{\tehole}}
           }
           {\dcasttwo{c}{b}{\tehole}}
\]
Consider the (indicated) function body,
%
where expansion inserts a cast on both the function expression~$x$ and its argument~$c$.
%
Together, these casts for~$x$ and~$c$ permit assigning a type to the
function body according to the rules in \Figref{fig:hasType}, where we
could not do so under the same context without casts.
%
We separately consider the expansions of~$x$ and of~$c$.

First, consider the function position of this application, here variable~$x$.
%
Without any cast, the type of variable~$x$ is hole type~$\tehole$;
however, the inserted cast on~$x$ permits treating it as though it has
arrow type $\tarr{\tehole}{\tehole}$, as described
in \Secref{sec:external-statics}.
%
The first three premises of Rule~{ESAp} accomplish this
%
by first synthesizing a type for the function expression, here
$\tehole$, then
%
by determining the matched arrow type~$\tarr{\tehole}{\tehole}$, and
finally,
%
by performing analytic expansion on the function expression with this
matched arrow type.
%
The resulting expansion has some type~$\tau_1'$ consistent with the matched arrow type.
%
In this case, because the subexpression~$x$ is a variable, analytic
expansion goes through subsumption so that type~$\tau_1'$ is
simply~$\tehole$.
%
The conclusion of the rule inserts the corresponding cast.
%
We go through type synthesis, \emph{then} analytic expansion, so that the hole
context records the matched arrow type for holes in function position,
rather than the type~$\tehole$ for all such holes, as would be the case
in a variant of this rule using synthetic expansion for the function
expression\todo{put this in the appendix?}{}.

Next, consider the application's argument, here constant~$c$.
%
The conclusion of Rule~{ESAp} inserts the cast on the argument's
expansion, from the type it is assigned by the final premise of the
rule~(type~$b$ here), to the argument type of the matched arrow type
of the function expression~(type~$\tehole$ here).

The example's second, outermost application uses through the same
application expansion rule.
%
In this case, the cast on the function is the identity cast for
$\tarr{\tehole}{\tehole}$.
%
For simplicity, we do not attempt to avoid the insertion of identity
casts in the core calculus; these will simply never fail during
evaluation.
%
However, it is safe to eliminate such identity casts during expansion,
and some formal accounts of gradual typing do so by defining three
application rules, including the original account of \cite{Siek06a}).

@@

Figures~\ref{fig:isGround}-\ref{fig:step} give dynamic semantics for \HazelnutLive,
%
resulting in a system capable of running incomplete programs without aborting when a hole is encountered.
%
We equip this semantics with a meaningful notion of type safety (involving both preservation and progress).
%
We also establish that the standard notion of type safety falls out when running complete terms.

The dynamic semantics consists of a ``small-step'' transition system,
%
where cast-related machinery closely follows the cast calculus from
the ``refined'' account of the gradually typed lambda calculus
by \citet{DBLP:conf/snapl/SiekVCB15}, which is known to be
theoretically well-behaved.
%
In particular, \Figref{fig:isGround} defines the judgement
$\isGround{\htau}$, which distinguishes the base type~$b$ and the
least specific arrow type~$\tarr{\tehole}{\tehole}$, as \emph{ground
types}; this judgement helps simplify the treatment of function casts.
%
\Figref{fig:isFinal} defines the judgement $\isFinal{d}$, which
distinguishes the final forms of the transition system.
%
There are two classes of final forms: (possibly-)boxed values and
indeterminate forms.%
%
\footnote{
        In most accounts of the cast calculus, values and ground types
        are distinguished with separate grammars together with an
        implicit identification convention.
        %
        Our judgmental formulation is more faithful to the mechanization and
        cleaner for our purposes, because we are distinguishing several
        classes of final forms.
}

@@

\subsubsection{Application and Substitution} 
%
Rule {ITBeta} in Fig.~\ref{fig:instruction-transitions} specifies the
standard beta reduction transition.
%
The bracketed premises of the form $\maybePremise{\isFinal{\dexp}}$ in
Fig.~\ref{fig:instruction-transitions}-\ref{fig:step} may be \emph{included}
to specify an eager, left-to-right evaluation strategy, or \emph{excluded} to
leave the evaluation order unspecified.
%
In our metatheory, we use the latter version of the rules, both for
the sake of generality, and to support fill-and-resume~(\Secref{sec:resumption}).

>>>>

The syntax of the core calculus specified in Fig.~\ref{fig:hazelnut-live-syntax} consists of types and expressions with holes. 
We distinguish between \emph{external} expressions, $e$, and \emph{internal} expressions, $d$. 
External expressions correspond to programs as entered by the programmer 
(see Sec.~\ref{sec:intro}\todo{or maybe Sec. 2?}{} for discussion of manual, semi-automated and fully automated hole entry methods). 
Each well-typed external expression (as specified in Sec.~\ref{sec:external-statics} below) expands to a well-typed internal expression (see Sec.~\ref{sec:expansion}) before it is evaluated (see Sec.~\ref{sec:evaluation}). 
We take this approach, notably also taken in the specification of Standard ML by \citet{Harper00atype-theoretic}, because (1) the external language supports type inference and explicit type ascriptions, $\hexp : \htau$, but it is formally simpler to eliminate ascriptions and specify a type assignment system when defining the dynamic semantics; and 
(2) we need additional syntactic machinery during evaluation for tracking hole closures and dynamic type casts. 
This machinery is inserted by the expansion step, rather than entered explicitly by the programmer. 
In this regard, the internal language is analagous to the cast calculus in the gradually typed lambda calculus \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will see the \HazelnutLive internal language goes beyond the cast calculus in several respects. We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently} 
(see Sec.~\ref{sec:agda-mechanization}). We have also implemented \HazelnutLive in a manner that maintains a novel end-to-end well-definedness invariant (see Sec.~\ref{sec:implementation}).

@@

We start with the type system of the \HazelnutLive
external language, which closely follows the \Hazelnut type
system \cite{popl-paper}; we discuss the minor differences as they come up below.

\Figref{fig:bidirectional-typing} defines the type system in the \emph{bidirectional} style 
%
with two mutually defined judgements \cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da}. The type synthesis
judgement~$\hsyn{\hGamma}{\hexp}{\htau}$ synthesizes a type~$\htau$
for external expression~$\hexp$ under typing context $\hGamma$, which tracks typing
assumptions of the form $x : \htau$ in the usual
manner \cite{pfpl,tapl}.
%
The type analysis judgement~$\hana{\hGamma}{\hexp}{\htau}$ checks
expression~$\hexp$ against a given type~$\htau$.
%
Algorithmically, analysis accepts a type as input, and synthesis gives
a type as output.
%
We start with synthesis for the programmer's ``top level'' external
exression.

@@

The primary benefit of specifying the \HazelnutLive external language 
bidirectionally is that the programmer does not need to annotate each hole with a type. 
%
An empty hole is
written $\hehole{u}$, where $u$ is a name for the hole, which we tacitly assume is unique. 
Holes did not have names in \Hazelnut. 
%
Rule {SEHole}\todo{rule name macros}{} specifies that an empty hole synthesizes hole type, written $\tehole$.
%
If an empty hole appears where an expression of some other type is
expected, e.g. under an explicit ascription (governed by Rule {SAsc})
or in the argument position of a function application (governed by
Rule {SAp}, discussed below), we apply the \emph{subsumption rule},
Rule {ASubsume}, which specifies that if an expression $e$ synthesizes
type $\htau$, then it may be checked against any \emph{consistent}
type, $\htau'$.

@@

For the familiar forms of the lambda calculus, the rules again follow prior work.
%
For simplicity, the core calculus includes only a single base type~$b$ with a single constant~$c$, governed by Rule {SConst} (i.e. $b$ is the unit type).
%
By contrast, \Hazelnut instead defines a number type with a single operation, which we include in Appendix~\ref{sec:extensions} alongside various other standard extensions to the core calculus\todo{do this, say more?}. 
%
Rule {SVar} specifies that variables synthesize the corresponding type from $\hGamma$.

@@

%The reason analytic expansion produces an expansion of consistent
%type is because the subsumption rule, as previously discussed, allows
%us to check an external expression against any type consistent with
%the type the expression actually synthesizes, whereas every internal
%expression can be assigned at most one type, i.e. the following
%standard unicity property holds of the type assignment system.
%
Analytic expansion produces an expansion of consistent type:
%
Inuitively, the subsumption rule, as previously discussed, allows us
to check an external expression against any type consistent with the
type the expression actually synthesizes.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% <<<<<<<<<<<<< Matt is here (latest pass over everything above is done -- 2018.04.11)

The rules governing expansion of constants, variables and lambda expressions --- Rules {ESConst}, {ESVar}, {ESLam} and {EALam} --- and the corresponding type assignment rules --- Rules {TAConst}, {TAVar} and {TALam} --- mirror the typing rules from Fig.~\ref{fig:bidirectional-typing} 
(so the corresponding cases of Theorem~\ref{thm:typed-expansion}, Theorem~\ref{thm:expandability} and Theorem~\ref{thm:expansion-generality} are straightforward). 
Note that in the internal language, all lambdas are half-annotated, again to support type assignment---Rule {EALam} inserts the annotation when expanding an unannotated external lambda based on the given type. 
The rules governing holes and the rules governing function applications and ascriptions, which both perform \emph{cast insertion}, are more interesting, so let us consider both cases in turn.

\subsubsection{Hole Expansion}\label{sec:hole-expansion} 
Rules {ESEHole}, {ESNEHole}, {EAEHole} and {EANEHole} govern the expansion of empty and non-empty expression holes to empty and non-empty \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$. 
%

The hole name, $u$, on a hole closure identifies the external hole that the hole closure corresponds to. 
Note that while we assume each hole name to be unique in the external language, there can be multiple hole closures with the same name during evaluation due to substitution. 
For example, the result from Fig.~\ref{fig:grades-example} showed four closures for the hole named 1\todo{update with actual number (four?) and hole name later}. 
There, we numbered each hole closure for a given hole sequentially, \li{1:1}, \li{1:2} and so on, but this is strictly for the sake of presentation so we omit hole closure numbers from the core calculus.

The hole expansion rules are the only rules that introduce hypotheses, of the form $\Dbinding{u}{\hGamma}{\htau}$, into the hole context, $\Delta$. 
The purpose of the hole context is to record a type, $\tau$, and a typing context, $\Gamma$, for each hole name, $u$.\footnote{
We use a hole context, rather than recording the typing context and type directly on each hole closure, to ensure that all closures for a hole name have the same typing context and type.} 
This notation for hole contexts is taken from contextual modal type theory (CMTT) \cite{Nanevski2008}, identifying hole names with metavariables and hole contexts with modal contexts (we say more about the connection with CMTT below). 
In all four hole expansion rules, the typing context recorded in the hole context is simply the current typing context when the hole is expanded. 
In the synthetic hole expansion rules, {ESEHole} and {ESNEHole}, the generated hole context assigns the hole type, $\tehole$, to $u$, as in the typing rules. 
However, the first two premises of the expansion subsumption rule, Rule EASubsume, disallow the use of subsumption for holes in analytic position. 
Instead, we have separate analytic rules, {EAEHole} and {EANEHole}, which record the type that the hole is being checked against into the hole context. 
This is again so that we can use type assignment for the internal language --- the type assignment rules TAEHole and TANEHole in Fig.~\ref{fig:hasType} assign a hole closure for $u$ the corresponding type from the hole context.

Each hole closure also has an associated environment, $\sigma$, which is a finite substitution of the form $[d_1/x_1, ~\cdots, d_n/x_n]$ for $n \geq 0$. 
The purpose of the closure environment is to keep a record of the substitutions that occur around the hole as evaluation occurs. 
Initially, no evaluation has occurred, so the initial environment generated by the hole expansion rules is the identity substitution for the typing context associated with $u$ in $\Delta$, which we notate $\idof{\hGamma}$ and define as follows.
\begin{defn}[Identity Substitution] $\idof{x_1 : \tau_1, ~\cdots, x_n : \tau_n} = [x_1/x_1, ~\cdots, x_n/x_n]$
\end{defn}
\noindent
The type assignment rules for hole closures, Rules TAEHole and TANEHole, require that we be able to check the environment of each hole closure against the corresponding typing context, written $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ and defined as follows:\todo{check that the definition in the Agda corresponds}
\begin{defn}[Substitution Typing]
$\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff $\domof{\sigma} = \domof{\hGamma'}$ and for each $x : \htau \in \hGamma'$ we have that $d/x \in \sigma$ and $\hasType{\Delta}{\hGamma}{d}{\tau}$.
\end{defn}
\noindent
It is easy to verify that the identity substitution satisfies this requirement, i.e. that $\hasType{\Delta}{\hGamma}{\idof{\hGamma}}{\hGamma}$.

Empty hole closures, $\dehole{u}{\sigma}{}$,  correspond to the metavariable closures (a.k.a. deferred substitutions) from CMTT, $\cmttclo{u}{\sigma}$.\todo{change notation from CMTT}{} 
We will see how closure environments evolve during evaluation in Sec.~\ref{sec:evaluation}. 
Non-empty hole closures, $\dhole{d}{u}{\sigma}{}$, do not directly correspond to a notion from CMTT (see Sec.~\ref{sec:resumption}).

\subsubsection{Cast Insertion}\label{sec:cast-insertion} 
%
Consider the following example: $\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$. 
The type synthesized for this example viewed as an external expression is $\tehole$, because the hole type annotation on $x$ allows us to apply it as a function of type $\tarr{\tehole}{\tehole}$, as previously discussed, and $c$ can be checked against type $\tehole$ by subsumption. 
However, viewed as an internal expression, this example is not well-typed---we do not have subsumption in the type assignment system defined in Fig.~\ref{fig:hasType}. 
Indeed, it would violate type safety if we could assign a type to this example in the internal language, because beta reduction of this example viewed as an internal expression would result in $c(c)$, which is clearly not well-typed.
The difficulty is that leaving the argument type unknown leaves how the argument is being used (in this case, as a function) also unknown.\footnote{In a system where type reconstruction is first used to try to fill in type holes, we could express a similar example by using $x$ at two or more different types, thereby causing type reconstruction to fail.
%On the other hand, if it is acceptable to arbitrarily choose one of the possible types, and type reconstruction is complete, then type holes will never appear in the internal language and the cast insertion machinery described in this section can be omitted entirely, leaving only the hole closure machinery described previously.
}
\todo{talk about elsewhere? maybe do type-hole-free version of calculus in appendix if there is time}{} 
By our  interpretation of hole types as unknown types from gradual type theory, we can address the problem by performing cast insertion. 
%

The cast form in Hazelnut Live is $\dcasttwo{\dexp}{\htau_1}{\htau_2}$, which, as specified by Rule TACast in Fig.~\ref{fig:hasType}, serves to ``box'' an expression of type $\htau_1$ for treatment as an expression of a consistent type $\htau_2$.%
\footnote{In the earliest work on gradual type theory, the cast form only gave the target type, $\htau_2$ \cite{Siek06a}, but it simplifies matters to include the assigned type, $\htau_1$, in the syntax \cite{DBLP:conf/snapl/SiekVCB15}.}

Casts are inserted during the expansion of function applications and ascriptions. 
The latter is more straightforward: 
Rule~{ESAsc} in Fig.~\ref{fig:expandSyn} inserts a cast from the assigned type to the ascribed  type. 
Theorem~\ref{thm:typed-expansion} inductively ensures that the two types are consistent.  
Note that we included ascription only for the sake of exposition---it can be defined using application together with the half-annotated identity, $e : \tau = \hap{(\halam{x}{\htau}{x})}{e}$, so application expansion, discussed next, is more general.

Cast insertion by the application expansion rule, Rule~{ESAp}, requires a bit more care. 
To understand the rule, consider the expansion for the example above:
\[\hap{\dcasttwo{(\halam{x}{\tehole}{\underline{\hap{\dcasttwo{x}{\tehole}{\tarr{\tehole}{\tehole}}}{\dcasttwo{c}{b}{\tehole}}}})}{\tarr{\tehole}{\tehole}}{\tarr{\tehole}{\tehole}}}{\dcasttwo{c}{b}{\tehole}}
\]
Let us focus on the expansion of the function body, underlined\todo{shaded?}, first. 
A cast has been inserted on both the function expression, $x$, and the the argument, $c$.

The cast on $x$ allows us to treat the variable $x$, which is of type $\tehole$, as being of the matched arrow type $\tarr{\tehole}{\tehole}$, as described in Sec.~\ref{sec:external-statics}. 
The first three premises of Rule~{ESAp} accomplish this by first synthesizing a type for the function expression, here $\tehole$, then determining the matched arrow type, $\tarr{\tehole}{\tehole}$, and then performing analytic expansion on the function expression with this matched arrow type.
The resulting expansion will have some type $\tau_1'$ consistent with the matched arrow type. 
In this case, because $x$ is of variable form, analytic expansion goes through subsumption so $\tau_1'$ is simply $\tehole$. 
The conclusion of the rule inserts the corresponding cast. 
We go through type synthesis then analytic expansion so that the hole context records the matched arrow type for holes in function position, rather than the type $\tehole$ for all such holes as would be the case in a variant of this rule using synthetic expansion for the function expression\todo{put this in the appendix?}{}.

The conclusion of Rule~{ESAp} inserts the cast on the argument's expansion, from the type it is assigned by the final premise of the rule, here $b$ as described at the start of Sec.~\ref{sec:expansion}, to the argument type of the matched arrow type of the function expression, here $\tehole$.

Observe that together, these casts allow us assign a type to the function body according to the rules in Fig.~\ref{fig:hasType}, where we could not do so under the same context without casts.

The outer application in the example above goes through the same rule. 
In this case, the cast on the function is the identity cast for $\tarr{\tehole}{\tehole}$. 
We do not attempt to avoid the insertion of identity casts in the core calculus for simplicity (these will simply never fail during evaluation), 
but it is safe to eliminate these during expansion (and some formal accounts of gradual typing do so by defining three application rules, including the original account of \cite{Siek06a}).

@@

To recap, the result of expansion is a well-typed internal expression with hole closures and casts, 
where the former corresponds to metavariable closures from CMTT \cite{Nanevski2008}, 
and the latter corresponds to casts from gradual type theory \cite{Siek06a,DBLP:conf/snapl/SiekVCB15}. 
However, the dynamic semantics for \HazelnutLive does not simply ``fall out'' from these observations.

The problem is first that \citet{Nanevski2008} defined only the logical reductions for CMTT, viewing it as a proof system for intuitionistic contextual modal logic via the propositions-as-types (Curry-Howard) principle. 
The paper therefore proved only a subject reduction property (which is closely related to type preservation). 
This is not a full dynamic semantics, and in particular, there is no notion of \emph{progress}, i.e. that well-typed terms cannot get ``stuck'' in an undefined state \cite{wright94:_type_soundness}. 
In any case, a conventional dynamic semantics for CMTT would not be immediately relevant to our goal of evaluating incomplete programs because, by our interpretation of hole closures, we would need a dynamic semantics for terms with free metavariables. 
\citet{Nanevski2008} sketched an interpretation of CMTT into the simply-typed lambda calculus with sums under permutation conversion%
\footnote{Permutation conversions are necessary to encode the commuting reductions of CMTT, which in turn are necessary to prove a strong normalization property. These issues are not relevant in \HazelnutLive because, as in the gradually typed lambda calculus, type holes admit non-termination: we can express the Y combinator as $(\halam{x}{\tehole}{x(x)}) (\halam{x}{\tehole}{x(x)})$.}, 
which has been studied by \citet{DBLP:journals/iandc/Groote02}, 
but under this interpretation an analagous problem arises---metavariables become variables of a function type, so again we cannot rely on the standard notion of progress on closed terms.\todo{citations}{}% We also cannot rely on, for example, weak head normalization because \HazelnutLive admits non-termination (due to casts).\todo{citation}

Furthermore, we need to integrate casts into the dynamic semantics. 
Fortunately, the dynamic semantics for the cast calculus from the gradually typed lambda calculus provides most, but not all, of the necessary machinery. 
The first problem is again with progress: in the cast calculus, the only irreducible terms of hole type are casts, which are accounted for by the progress theorem, but in $\HazelnutLive$, holes induce additional irreducible terms. 
The second missing piece is that in prior work on casts, evaluation aborts when cast failure occurs. 
Our goal, as discussed in Sec.~\ref{sec:failed-cast-example}\todo{where?}{}, is for cast failure to instead insert a membrane around the dynamic type error, 
much like a non-empty hole serves as a membrane around a static type error, 
allowing in both cases for evaluation to safely and meaningfully continue past the error when desired.

The dynamic semantics for \HazelnutLive specified in Figures~\ref{fig:isGround}-\ref{fig:step} addresses the difficulties just outlined, 
resulting in a system capable of running incomplete programs without aborting when a hole is encountered. 
This semantics is equipped with a meaningful notion of type safety (involving both preservation and progress). 
We also establish that the standard notion of type safety falls out when running complete terms.

The dynamic semantics is specified as a ``small-step'' transition system. 
The cast-related machinery is based closely on the cast calculus from the ``refined'' account of the gradually typed lambda calculus by \citet{DBLP:conf/snapl/SiekVCB15}, which is known to be theoretically well-behaved. 
In particular, Fig.~\ref{fig:isGround} defines the judgement $\isGround{\htau}$, which distinguishes the base type, $b$, and the least specific arrow type, $\tarr{\tehole}{\tehole}$, as \emph{ground types}, which play a role in simplifying the treatment of function casts. 
Fig.~\ref{fig:isFinal} defines the judgement $\isFinal{d}$, which distinguishes the final forms of the transition system. 
There are two classes of final forms: (possibly-)boxed values and indeterminate forms.%
\footnote{In most accounts of the cast calculus, values and ground types are distinguished with separate grammars together with an implicit identification convention. 
Our judgemental formulation is more faithful to the mechanization and cleaner for our purposes, because we are distinguishing several classes of final forms.}

@@

\subsubsection{Application and Substitution} 
Rule {ITBeta} in Fig.~\ref{fig:instruction-transitions} specifies the standard beta reduction transition. 
The bracketed premises of the form $\maybePremise{\isFinal{\dexp}}$ in Fig.~\ref{fig:instruction-transitions}-\ref{fig:step} can be included to specify an eager, left-to-right evaluation strategy, 
or excluded to leave the evaluation order unspecified. 
We do the latter in our metatheory, both for the sake of generality and for reasons we return to in Sec.~\ref{sec:resumption}.
